\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-07 paper/algorithm.tex
paper: rewrite algorithm section for clarity

\subsection{Search space}
\label{sec:search-space}
To reduce the size of the query plan search space, only plans that fit the following template are considered. First, all filters are pushed to the leaves of the query tree, immediately after the scans. Joins are performed after filtering, and only left-deep plans are considered. Any group-by/aggregate will be performed after the final join. Finally, projections are placed at the root of the query tree. The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.

@@

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan is associated with a set of dirty attributes $D$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ may contain null. We compute a dirty set for each base table using the table statistics, which track the number of null values in each column. If we apply an imputation operator to a relation $R$ with a dirty set $D$, $\mu_C (R)$ or $\delta_C (R)$, the resulting query has a dirty set $D' = D \setminus C$.  Applying a projection $\pi_C(R)$ produces a dirty set $D' = D \cap C$. A join $R_1 \Join_\psi R_2$ with dirty sets $D_1$ and $D_2$  produces a dirty set $D' = D_1 \cup D_2$.  Filters do not change the dirty set.

That is, the dirty set over-approximates the set of attributes that contain null. For example, a filter might remove all tuples which contain null without changing the dirty set, forcing an unnecessary imputation. We choose to over-approximate the dirty set to avoid the possibility of dropping a tuple that contains a null value without explicitly imputing the value or applying a drop operator.

>>>>

\subsection{Search space}
\label{sec:search-space}
To reduce the size of the query plan search space, only plans that fit the following template are considered.

\begin{itemize}
\item All filters are pushed to the leaves of the query tree, immediately after scanning a table. We use a pre-processing step which conjoins together filters which operate on the same table, so we can assume that each table has at most one relevant filter.

\item Joins are performed after filtering, and only left-deep plans are considered.

\item A grouping \& aggregation operator is optional, and if present will be performed after the final join.

\item Projections are placed at the root of the query tree.
\end{itemize}

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before/after traditional operators.

@@

Imputation operators can be placed at any point in the query plan, but to meet the guarantee that no non-imputation operator sees a null value, there are cases where an imputation operator is required. To track these cases, each query plan $q$ is associated with a set of dirty attributes $\textsc{Dirty}(q)$. An attribute $c$ is \emph{dirty} in some relation if the values for $c$ contain null. We compute a dirty set for each table using the table statistics, which track the number of null values in each column. The dirty set for a query plan can be computed recursively as follows:

\begin{align*}
  \textsc{Dirty}(q) = \begin{cases}
    \textsc{Dirty}(q') \setminus C & q = \delta_C(q')\ \text{or}\ q = \mu_C(q') \\
    \textsc{Dirty}(q') \cap C & q = \pi_C(q') \\
    \textsc{Dirty}(q_l) \cup \textsc{Dirty}(q_r) & q = q_l \bowtie_\psi q_r \\
    \textsc{Dirty}(q') & q = \sigma_\phi(q') \\
    \textsc{Dirty}(t) & q = \text{some table}\ t \\
  \end{cases}
\end{align*}

Note that \textsc{Dirty} over-approximates the set of attributes that contain null. For example, a filter might remove all tuples which contain null, but the dirty set would be unchanged. We choose to over-approximate to ensure that all null values are explicitly imputed or dropped.

### 2017-02-07 paper/algorithms/plan-helpers.tex
paper: rewrite algorithm section for clarity

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan, $C_{req}$ is a set of attributes that must be imputed.}
    \Ensure{Returns a set of query plans such that $\Call{Dirty}{q'} \cap C_{req} = \emptyset$.}
    \Function{AddImpute}{$q, C_{req}$}
    \State $C_{min} \gets \Call{Dirty}{q} \cap C_{req}$
    \If{$\Call{Dirty}{q} = \emptyset$}
    	\State \Return $\{ q \}$
   \ElsIf{$C_{min} = \emptyset$}
    	\State \Return $\{ \mu_{\Call{Dirty}{q}}(q), q \}$
    \Else
    \State \Return $\{\mu_{\Call{Dirty}{q}}(q), \mu_{C_{min}}(q), \delta_{C_{min}}(q)\}$
    \EndIf
    \EndFunction

@@

    \Require{$Q$ is a set of query plans.}
    \Ensure{Returns an optimal query plan for each distinct dirty set in $Q$.}
    \Function{OptRel}{$Q$}
    \State $D \gets \{\Call{Dirty}{q} ~|~ q \in Q\}$
    \State \Return $\{\argmin_{q \in Q \land \Call{Dirty}{q} = d} \Call{Cost}{q} ~|~ d \in D\}$
    \EndFunction

@@

    \Require{$t$ is a table and $\phi$ is a filter predicate.}
    \Ensure{Returns a set of optimal query plans for scanning and filtering $t$, with distinct dirty sets.}
    \Function{OptFilter}{$t, \phi$}
    \State \Return $\Call{OptRel}{\{  \sigma_{\phi}(q) | q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}} \} }$
    \EndFunction

@@

    \Require{$Q$ is a map from tables and dirty sets to optimal query plans (possibly with imputations), and $\Psi$ relates query plans with join predicates.}
    \Ensure{Returns a map from dirty sets to optimal query plans involving all necessary joins}
    \Function{OptJoin}{$Q, \Psi$}
    \For{$size \in 1...|\Psi|$}
    	\State $S \gets subset(\Psi, size)$
	\For{$\psi \in S$}
		\State $S' \gets S \setminus \psi$
			\If{$joins(S', \psi)$} 
				\State $S'_{plans} \gets Q(\Call{Rels}{S'})$
				\State $t \gets \Call{Rels}{\psi} \setminus \Call{Rels}{S'}$
				\State $t_{plans} \gets Q(t)$
				\For {$l, r \in S'_{plans} \times t_{plans}$}
					\State $P_l \gets \Call{AddImpute}{l, \Call{Attrs}{\psi}}$
					\State $P_r \gets \Call{AddImpute}{r, \Call{Attrs}{\psi}}$
					\State \Call{Update}{Q, $\Call{OptRel}{ \{p_l \bowtie_{\psi} p_r | p_l \in P_l, p_r \in P_r \} }$}
				\EndFor
		\EndIf
	\EndFor
    \EndFor
    \State \Return{Q(\Call{Rels}{$\Phi$})}
    \EndFunction
  \end{algorithmic}
  \caption{Base algorithms for query planning with imputations.}
  \label{algo:plan-helpers}
%\end{algorithm}

>>>>

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan. $C_{req}$ is the set of attributes that must be imputed in the output of this query plan. $C_{vis}$ is the set of attributes which are visible in the output of the overall query.}
    \Ensure{A set of query plans $Q$ with added imputation.}
    \Function{AddImpute}{$q, C_{req}, C_{vis}$}
    \State $D_{must} \gets \Call{Dirty}{q} \cap C_{req}$
    \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_{vis})$
    \State $Q \gets \emptyset$
    \If{$D_{must} = \emptyset$}
    \State $Q \gets Q \cup \{q\}$
    \Else
    \State $Q \gets Q \cup \{\mu_{D_{must}}(q),\ \delta_{D_{must}}(q)\}$
    \EndIf

    \If{$D_{may} \neq \emptyset$}
    \State $Q \gets Q \cup \{\mu_{D_{may}}(q)\}$
    \EndIf

    \State \Return $Q$
    \EndFunction

@@

    \Require{$t$ is a table and $\phi$ is a filter predicate.}
    \Ensure{A set of plans for filtering $t$ with $\phi$.}
    \Function{OptFilter}{$t, \phi, C_{vis}$}
    \State \Return $\{\sigma_{\phi}(q)\ |\ q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}, C_{vis}}\}$
    \EndFunction

@@

    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, and $C_{vis}$ is a set of visible attributes.}
    \Ensure{A set of plans for grouping \& aggregating with $G$ and $A$.}
    \Function{OptGroupBy}{$q, G, A, C_{vis}$}
    \State $C \gets G \cup \Call{Attrs}{A}$
    \State $Q' \gets \Call{AddImpute}{q, C, C_{vis}}$
    \State \Return $\{\Call{GroupBy}{q', G, A} ~|~ q' \in Q'\}$
    \EndFunction

@@

    \Require{$q$ is a query plan, $P$ is a set of projection attributes.}
    \Ensure{A set of plans with the attributes in $P$ projected.}
    \Function{OptProject}{$q, P$}
    \State \Return $\{\pi_P(q') ~|~ q' \in \Call{AddImpute}{q, P, P}\}$
    \EndFunction

    \Require{A set of tables $T$, a plan cache $Q$ containing at least one query plan for each table in $T$, and a relation $J : T \times \Psi \times T$ between tables and join predicates.}
    \Ensure{An updated plan cache $Q'$ which contains at least one plan that combines all of the tables in $T$.}
    \Function{OptJoin}{$Q, T, J, C_{vis}$}
    \For{$size \in 2...|T|$}
    \For{$S \in \{\text{all length}\ size\ \text{subsets of}\ T\}$}
    \For{$t \in S$}
    \State $S' \gets S \setminus \{t\}$
    \For{$(t, \psi, t') \in J$ where $t' \in S'$}
    \State $C \gets \Call{Attrs}{\psi}$
    \State $L \gets \{\Call{AddImpute}{q, C, C_{vis}} ~|~ q \in Q[t] \}$
    \State $R \gets \{\Call{AddImpute}{q, C, C_{vis}} ~|~ q \in Q[S'] \}$
    \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L, r \in R\}$
    \EndFor
    \EndFor
    \EndFor
    \EndFor
    \EndFunction
  \end{algorithmic}
  \caption{Base algorithms for query planning with imputations.}
  \label{algo:plan-helpers}
%\end{algorithm}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:

### 2017-02-07 paper/algorithms/plan-toplevel.tex
paper fix: extract algorithm begin/end markers to be able to use with [H] placement in slides
paper: rewrite algorithm section for clarity

\begin{algorithm}
 \begin{algorithmic}
   \Require{set of tables $T$, filter predicates $\Phi$, join predicates $\Psi$, a set of projection attributes $P$, optional set of grouping attributes $G$ and aggregator function $A$}
   \Ensure{Returns optimized query plan.}

  \Function{Plan}{$T, \Phi, \Psi, P, G, A$}
 \State  $Q \gets \{ \}$ \Comment{map from table and dirty set to plans }
   \For{$t \in T$}
   	\State $\phi \gets \Call{GetFilters}{t, \Phi}$
	\State $P_{t} \gets \Call{OptFilter}{t, \phi}$
	\State $Q \gets Q \cup \{ (t, \Call{Dirty}{q}) \mapsto q | q \in P_t \}$
   \EndFor

   \State $M \gets \{ \}$ \Comment{map from dirty set to plans, contains all joins necessary}
   \State $M \gets  \Call{OptJoin}{Q, \Psi}$	\Comment{optimize joins}
   \If{$G \neq \emptyset \land A \neq NULL$}
   	\State $attrs \gets \Call{Attrs}{G} \cup \Call{Attrs}{A}$
   	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \Call{GroupBy}{\Call{AddImpute}{q, attrs}, G, A}}$
   \Else
  	\State $M \gets \Call{OptRel}{\bigcup_{q \in M} \pi_{P}(\Call{AddImpute}{q, P}) }$
  \EndIf

\Return $\argmin_{q \in M} Cost(q)$
\EndFunction

  \end{algorithmic}
  \caption{Top-level query planner with imputations.}
\label{algo:top-level-planner}
\end{algorithm}

>>>>

% \begin{algorithm}
\begin{algorithmic}
  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$, and an optional set of grouping attributes $G$ and aggregator function $A$.}
  \Ensure{An optimized query plan.}

  \Function{Plan}{$T, F, J, P, G, A$}
  \State Let $Q$ be an empty plan cache.
  \For{$t \in T$} \Comment{Add selections to the plan cache.}
  \If {$\exists \phi: (t, \phi) \in F$}
  \State $Q[\{t\}] \lhd \Call{OptFilter}{Q, t, \phi, P}$
  \Else
  \State $Q[\{t\}] \lhd \{t\}$
  \EndIf
  \EndFor

  \State $\Call{OptJoin}{Q, T, J, P}$	\Comment{Optimize joins.}
  \State $B \gets Q[T]$ \Comment{Get the best plans for all tables.}
  \If{$G \neq \emptyset$} \Comment{Add optional group \& aggregate.}
  \State $B \gets \bigcup_{q \in B} \Call{OptGroupBy}{q, G, A, C_{vis}}$
  \EndIf

  \State $B \gets \bigcup_{q \in B} \Call{OptProject}{q, P}$ \Comment{Add projections.}

  \State \Return $\argmin_{q \in B} Cost(q)$ \Comment{Return cheapest plan.}
  \EndFunction

\end{algorithmic}
\caption{Top-level query planner with imputations.}
\label{algo:top-level-planner}
% \end{algorithm}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:
